miniKanren with fair search strategies

### 2019-05-20 Paper/paper.tex
bracket to paranthesis

The syntax of a programming language should reflect its semantics. When writing 
a \conde{} expression in miniKanren, a programmer would expect all clauses 
share the same chance of being explored, as these clauses are written in 
parallel. The existing search strategy, interleaving depth-first search 
(\DFSi{}), 
however, prioritize its clauses by the order how they are written down. 
Similarly, when a \conde{} is followed by another goal conjunctively, a 
programmer would expect answers in parallel share the same chance of being 
explored. Again, the answers by \DFSi{} is different from the 
expectation. We have devised three new search strategies that have different 
level of fairness in \disj{} and \conj{}.

@@

% \conde is the flagship disjunction operator in miniKanren.
A subtlety arises when a \conde contains many clauses: not every clause has an 
equal chance to contribute to the result. As an example, consider the following 
relation \texttt{repeato} and its invocation.

@@

\conde intuitively relates its clauses with logical \texttt{or}. And thus an 
unsuspicious beginner would expect each letter to contribute equally to the 
result, as follows.

@@

The \conde in TR2, however, generates a less expected result.

@@

The specification that every clause in the same \conde{} is given equal 
``search priority'' is called fair \disj{}. And search strategies with 
almost-fair \disj{} give every clause in the same \conde{} similar priority. 
Fair \conj{}, a related concept, is more complicated. We defer it to the next 
section.

To summarize our contribution, we
\begin{itemize}
	\item propose and implement balanced interleaving depth-first search 
	(\DFSbi{}), a new search strategy with almost-fair \disj{}.
	\item propose and implement fair depth-first search (\DFSf{}), a 
	new search strategy with fair \disj{}.
	\item implement in a new way breath-first search (BFS), a search strategy 
	with fair \disj{} and fair \conj{}. (our code runs faster in all 
benchmarks and is simpler). And we prove formally that our BFS implementation 
is equivalent to the one by \citet{seres1999algebra}.
\end{itemize}

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the answer. \DFSi, the current search
strategy, however, results in many more lists of \texttt{a}s than lists
of other letters. And some letters  (e.g. \texttt{c} and \texttt{d}) are
rarely seen. The situation would be exacerbated if \conde contains more clauses.

@@

Under the hood, the \conde here is allocating computational resource to 
four trivially different search space. The unfair \disj{} in 
\DFSi{} allocates many more resources to the first search space. On the 
contrary, fair \disj{} would allocate resources evenly to each search space.

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the answer. Search strategies with unfair \conj{} (e.g. 
\DFSi, \DFSbi, \DFSf), however, results in many more lists of \texttt{a}s than 
lists of other letters. And some letters are rarely seen. The situation would 
be exacerbated if \conde contains more clauses.

@@

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-conj-fair.rkt}
	\end{tabular}
\end{center}

@@

\begin{figure}
	\lstinputlisting{Figures/balanced-disj.rkt}
	\caption{implementation of DFS$_\textrm{bi}$}
	\label{balanced-disj}
\end{figure}

@@

\begin{figure}
	\lstinputlisting{Figures/fDFS.rkt}
	\caption{implementation of DFS$_\textrm{f}$}
	\label{fDFS}
\end{figure}

@@

In this section, we compare the efficiency of search strategies. A concise 
description is in Table~\ref{compare-efficiency}. A hyphen means running out of 
memory. The first two benchmarks are taken from 
\citet{friedman_reasoned_2018}. \texttt{reverso} is from 
\citet{rozplokhas2018improving}. Next two benchmarks 
about quine are modified from a similar test case in \citet{byrd2017unified}. 
The modifications are made 
to circumvent the need for symbolic constraints (e.g. $\neq$, 
\texttt{absento}). Our version generates de 
Bruijnized expressions and prevent closures getting into list. The two 
benchmarks differ in the \conde clause order of their relational interpreters. 
The last two 
benchmarks are about synthesizing expressions that evaluate to \texttt{'(I love 
you)}. This benchmark is also inspired by \citet{byrd2017unified}. Again, the 
sibling benchmarks differ in the \conde clause order of their relational 
interpreters. The first one 
has elimination rules (i.e. application, \texttt{car}, and \texttt{cdr}) at the 
end, while the other has them at the beginning. We conjecture that DFS$_\textrm{i}$ would 
perform badly in the second case because elimination rules complicate the 
problem when running backward. The evaluation supports our conjecture.

>>>>

The syntax of a programming language should reflect its semantics. When writing 
a \conde{} expression in miniKanren, a programmer would expect all clauses 
share the same chance of being explored, as these clauses are written in 
parallel. The existing search strategy, interleaving depth-first search 
(\DFSi{}), 
however, prioritizes its clauses by the order how they are written down. 
Similarly, when a \conde{} is followed by another goal conjunctively, a 
programmer would expect states in parallel share the same chance of being 
explored. Again, the answers by \DFSi{} is different from the 
expectation. We have devised three new search strategies that have different 
level of fairness in \disj{} and \conj{}.

@@

\NOTE[LKC]{Shall we delete the sentence about relational interpreters? It is 
misleading: it is emphasizing relational interpreters, but we say nothing 
further about them later.}

% \conde{} is the flagship disjunction operator in miniKanren.
A subtlety arises 
when a \conde{} contains many clauses: not every clause has an 
equal chance to contribute to the result. As an example, consider the following 
relation \texttt{repeato} and its invocation.

@@

\conde{} intuitively relates its clauses with logical \texttt{or}. And thus an 
unsuspicious beginner would expect each letter to contribute equally to the 
result, as follows.

@@

The \conde{} in TR2, however, generates a less expected result.

@@

The specification that every clause in the same \conde{} is given equal 
``search priority'' is called fair \disj{}. And search strategies with 
almost-fair \disj{} give every clause similar priority. 
Fair \conj{}, a related concept, is more complicated. We defer it to the next 
section.

To summarize our contribution, we
\begin{itemize}
	\item propose and implement balanced interleaving depth-first search 
	(\DFSbi{}), a new search strategy with almost-fair \disj{}.
	\item propose and implement fair depth-first search (\DFSf{}), a 
	new search strategy with fair \disj{}.
	\item implement in a new way breath-first search (BFS), a search strategy 
	with fair \disj{} and fair \conj{}. And we prove formally that our BFS 
implementation is equivalent to the one by \citet{seres1999algebra}. Our code 
runs faster in all benchmarks and is simpler. 
\end{itemize}

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the query result. \DFSi, the current search
strategy, however, results in many more lists of \texttt{a}s than lists
of other letters. And some letters  (e.g. \texttt{c} and \texttt{d}) are
rarely seen. The situation would be exacerbated if \conde{} contains more 
clauses.

@@

Under the hood, the \conde{} here is allocating computational resource to 
four trivially different search space. The unfair \disj{} in 
\DFSi{} allocates many more resources to the first search space. On the 
contrary, fair \disj{} would allocate resources evenly to each search space.

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the answer. Search strategies with unfair \conj{} (e.g. 
\DFSi, \DFSbi, \DFSf), however, results in many more lists of \texttt{a}s than 
lists of other letters. And some letters are rarely seen. The situation would 
be exacerbated if \conde{} contains more clauses.

@@

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-conj-BFS.rkt}
	\end{tabular}
\end{center}

@@

\begin{figure}
	\lstinputlisting{Figures/DFSbi.rkt}
	\caption{implementation of \DFSbi{}}
	\label{balanced-disj}
\end{figure}

@@

\begin{figure}
	\lstinputlisting{Figures/DFSf.rkt}
	\caption{implementation of \DFSf{}}
	\label{fDFS}
\end{figure}

@@

In this section, we compare the efficiency of search strategies. A concise 
description is in Table~\ref{compare-efficiency}. A hyphen means running out of 
memory. The first two benchmarks are taken from 
\citet{friedman_reasoned_2018}. \texttt{reverso} is from 
\citet{rozplokhas2018improving}. Next two benchmarks 
about quine are modified from a similar test case in \citet{byrd2017unified}. 
The modifications are made 
to circumvent the need for symbolic constraints (e.g. $\neq$, 
\texttt{absento}). Our version generates de 
Bruijnized expressions and prevent closures getting into list. The two 
benchmarks differ in the \conde{} clause order of their relational 
interpreters. 
The last two 
benchmarks are about synthesizing expressions that evaluate to \texttt{'(I love 
you)}. This benchmark is also inspired by \citet{byrd2017unified}. Again, the 
sibling benchmarks differ in the \conde{} clause order of their relational 
interpreters. The first one 
has elimination rules (i.e. application, \texttt{car}, and \texttt{cdr}) at the 
end, while the other has them at the beginning. We conjecture that DFS$_\textrm{i}$ would 
perform badly in the second case because elimination rules complicate the 
problem when running backward. The evaluation supports our conjecture.
